---
title: "Predictive Eviction Prevention"
subtitle: "CPLN 5090 Final, Technical Appendix"
author: "Mackenna Amole, Ethan Harner, Angie Kwon"
format: 
  html:
    toc: true
    toc-location: left
    code-fold: show
    theme: cosmo
editor: 
  markdown: 
    wrap: 72
---

# Executive Summary

This technical appendix documents the workflow from exploratory data
analysis through model evaluation and policy implications of creating a
model to predict patterns of eviction across Philadelphia census tracts.
Eviction filings have seen fluctuation since the pandemic, but patterns
remain with activity concentrated in North and West Philadelphia. This
analysis helps to understand the spatial and demographic factors
connected to evictions, and then proposing recommendations for policy to
address these issues. Our final model identifies higher-risk census
tracts where proactive eviction prevention measures and policies should
be allocated. The patterns identified by the model can support
decision-making around renter assistance and resource distribution to
proactively target areas with disproportionate patterns of eviction.

# Introduction and Problem Statement

Following the COVID-19 Pandemic, evictions have dropped across
Philadelphia. This is a positive trend, and is reflects the positive
impact that anti-eviction city policy has had following nation-wide
eviction moratoriums placed in effect in early 2020. Six years on from
initial outbreak, average eviction counts by census tract have fallen
city-wide; and have seemingly stabilized since 2022.

However, despite these positive trends, certain neighborhoods across the
city still experience high eviction counts - with many high
eviction-count census tracts clustered in North and West Philadelphia.
Given that high prevalence of evictions have become spatially clustered,
we seek to develop a predictive model that is capable of predicting
which census tracts will have a higher than average eviction count.

This model's spatial component is particularly important for
understanding efficient resource allocation, as census tracts have
different average rents, and thus emergency rental assistance
allocations would need to be adjusted accordingly. Furthermore, a
spatially-predictive prevention model enables the deployment of on-the
ground resources such as housing counseling and homelessness services.
Finally, knowing the spatial distribution of future evictions, allows
city officials to proactively look for patterns in housing quality,
landlord type, and other spatial data which connects to eviction
likelihood.

#Exploratory Data Analysis

```{r}
#| code-fold: true
#| code-summary: "Required Packages"
#| message: false
#load in packages
library(tidyverse)
library(dplyr)
library(lubridate)
library(sf)
library(tigris)
library(tidycensus)
library(scales)
library(viridis) 
library(knitr)
library(kableExtra)
library(patchwork)
library(zoo)
library(stringr)
library(ggplot2)
library(here)
library(tidymodels)
options(tigris_use_cache = TRUE,
        tigris_progress = FALSE)
```

Before starting our exploratory analysis, we cleaned the data to remove
null value in date and location data, and replace null values in
eviction counts with zeros. We then began to construct exploratory data
sets that visualize city-wide eviction filings over time. We then
constructed spatial data sets through gathering census shape files and
joining them with the eviction data set. Evictions were already
aggregated to a census-tract level.

```{r data cleaning}
#| code-fold: true
#| code-summary: "Cleaning Eviction Filing Data"
#| message: false
# Load in data
eviction_raw <- read.csv(here("data/philadelphia_monthly_2020_2021.csv"))

eviction <- eviction_raw %>%
  drop_na(
    type,
    GEOID,
    month,
  ) %>%
  filter(
    GEOID != "sealed"
  ) %>%
  mutate(
    year = str_sub(month, 4,7),
    month = str_sub(month, 1,2),
    filings_2020 = ifelse(is.na(filings_2020), 0, filings_2020),
    filings_avg = ifelse(is.na(filings_avg), 0, filings_avg),
    filings_avg_prepandemic_baseline = ifelse(is.na(filings_avg_prepandemic_baseline), 0, filings_avg_prepandemic_baseline)
  ) %>%
  relocate(year, .after = month)

#glimpse(eviction)

# How many total evictions?
cat("Total evictions 2020-2025:", nrow(eviction), "\n")

# Date range
cat("Date range:", 
    min(eviction$month),"/",min(eviction$year), "to", 
    max(eviction$month),"/",max(eviction$year), "\n")

# Summary per month per year
evictions_summary <- eviction %>%
  st_drop_geometry() %>%
  group_by(GEOID, month, year) %>%
  summarize(
    total_filings = sum(filings_2020, na.rm = TRUE),
    filings_avg = mean(filings_avg, na.rm = TRUE),
    prepan_avg = mean(filings_avg_prepandemic_baseline, na.rm = TRUE),
    racial_majority = racial_majority
  )

#Evictions Filings at a City Level Over Time
evictions_over_time_city_wide <- eviction %>%
  st_drop_geometry() %>%
  group_by(month, year) %>%
  summarize(
    total_filings = sum(filings_2020, na.rm = TRUE),
    filings_avg = mean(filings_avg, na.rm = TRUE),
  )

# Census tracts for when we need them
tracts_phl <- tracts(state = "PA", county = "Philadelphia", year = 2025) %>%
  st_transform(4326)

# Joining evictions data with census data and census geometry - allowing us to visualize evictions spatially
spatial_evictions <- eviction %>%
  left_join(tracts_phl, by = "GEOID")

spatial_evictions <- st_as_sf(spatial_evictions)
```

## Trends in Post-Pandemic Eviction Filings

From our exploratory analysis, we see that evictions fell sharply
following the beginning of the pandemic - a response to eviction
moratorium placed into effect following lock-down orders. Evictions
slowly increased city-wide, peaking in late summer of 2022. This is
possibly due to COVID-19 emergency rental assistance relief money drying
up in late 2022, leaving some renters unable to pay accumulated arrears
or unable to cover lapses in income. Evictions stabilized through the
fall and winter of 2022, and have relatively plateaued at approximately
1,200 filings since 2023. Notably though, evictions have not increased
to a February 2022 baseline since the pandemic.

```{r visualizing city-wide eviction trends}
#| code-fold: true
#| code-summary: "Cleaning Eviction Filing Data"
#| message: false
evictions_over_time_city_wide <- evictions_over_time_city_wide %>%
  mutate(date = as.Date(paste(year, month, "01", sep = "-")))

ggplot(evictions_over_time_city_wide, aes(x = date)) +
  geom_col(aes(y = total_filings, fill = "Total filings"),
           alpha = 0.6, width = 25) +
  scale_x_date(
    date_breaks = "3 months",
    date_labels = "%b '%y",
    expand = expansion(mult = 0.01)
  ) +
  scale_y_continuous(
    labels = comma,
    expand = expansion(mult = c(0, 0.05))
  ) +
  scale_fill_manual(values = c("Total filings" = "#378ADD")) +
  labs(
    title = "Eviction filings over time — city wide",
    subtitle = "Monthly total filings",
    x = NULL,
    y = "Filings",
    fill = NULL,
    color = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(color = "gray40", size = 11),
    legend.position = "top",
    legend.justification = "left",
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

Examining where the most evictions have taken place city-wide, we can
see that since evictions began to return to normal in 2022, the majority
of evictions have taken place in North and West Philadelphia. This
largely makes sense given that these neighborhoods are lower income and
more renter dominated compared to other parts of the city. However,
there are hot spots of evictions in the far northeast and northwest that
bare further consideration, as they are seemingly not characteristic of
the other neighborhoods identified as being hotspots of evictions.

```{r visualizing city-wide eviction patterns over time}
#| code-fold: true
#| code-summary: "Spatial Distribution of Filings Over Time"
#| message: false
eviction_map_data <- spatial_evictions %>%
  filter(year >= 2020 & year <= 2026) %>%
  group_by(GEOID, year) %>%
  summarise(
    total_filings = sum(filings_2020, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(eviction_map_data) +
  geom_sf(aes(fill = total_filings), color = NA) +
  scale_fill_viridis_c(
    option = "magma",
    direction = -1,
    name = "Total filings",
    labels = scales::comma,
    na.value = "grey85"
  ) +
  facet_wrap(~ year, ncol = 4) +
  labs(
    title = "Eviction filings by census tract",
    subtitle = "Annual totals, 2020–2026, NOTE: 2026 data incomplete",
    caption = "Source: Eviction Lab"
  ) +
  theme_void(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold", size = 14, margin = margin(b = 4)),
    plot.subtitle = element_text(color = "gray40", size = 11, margin = margin(b = 12)),
    plot.caption = element_text(color = "gray60", size = 9),
    legend.position = "bottom",
    legend.key.width = unit(2, "cm"),
    legend.key.height = unit(0.3, "cm"),
    legend.title = element_text(size = 10),
    strip.text = element_text(face = "bold", size = 11, margin = margin(b = 6)),
    panel.spacing = unit(0.5, "cm"),
    plot.margin = margin(12, 12, 12, 12)
  )
```

Most neighborhoods dropped in alignment with the city's declining
eviction trend. Most notably, southwest and north Philadelphia see very
strong drops in evictions relative to pre-pandemic data. However,
neighborhoods in the river wards and northwest have increased in average
evictions. This presents an area for further exploration, understanding
and predicting why certain neighborhoods are not taking advantage of
post-pandemic low evictions.

```{r change in evictions post pandemic}
#| code-fold: true
#| code-summary: "Change Since Pandemic Baseline"
#| message: false
evictions_change <- spatial_evictions %>%
  filter(year == "2025" & month == "12") %>% #oddly it seems that the pandemic baseline numbers haven't been filled in for 2026 data? We should discuss this. So as of now, I've selected for december 2025.
  mutate(post_pandemic_change = filings_avg - filings_avg_prepandemic_baseline)

ggplot(evictions_change) +
  geom_sf(aes(fill = post_pandemic_change), color = NA) +
  scale_fill_gradient2(
    low = "#2166ac",
    mid = "white",
    high = "#d6604d",
    midpoint = 0,
    name = "Change in filings\nvs. pre-pandemic avg",
    labels = scales::comma,
    na.value = "grey85"
  ) +
  labs(
    title = "Change in eviction filings vs. pre-pandemic baseline",
    subtitle = "December 2025, census tract level",
    caption = "Positive values (red) = more filings than pre-pandemic average"
  ) +
  theme_void(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold", size = 14, margin = margin(b = 4)),
    plot.subtitle = element_text(color = "gray40", size = 11, margin = margin(b = 12)),
    plot.caption = element_text(color = "gray60", size = 9),
    legend.position = "right",
    legend.key.width = unit(0.4, "cm"),
    legend.key.height = unit(1.5, "cm"),
    plot.margin = margin(12, 12, 12, 12)
  )

```

## Included Variables and Rationale

In predicting evictions, we relied heavily on demographic and housing
data as key predictors of eviction likelihood. An outline of our
included variables is below, alongside justifications and limitations of
each.

1.  **Housing Costs:**

    We included housing costs under the rationale that higher costs
    would produce higher barriers to timely payment of rent, ultimately
    producing evictions. Notably, this might also capture high-end
    markets, but we believe that its inclusion is justified, as cost
    burdened renters with higher gross levels of rent are less likely to
    be able to access additional financial resources sufficient to cover
    gaps in rent compared to renters in lower cost markets.

2.  **Rent Burden**

    Rent burden is included under the straightforward assumption that
    higher cost burden will produce higher risks of non-payment.
    Non-payment may lead to evictions.

3.  **Higher Education Attainment**

    Higher education levels are expected to be correlated with increased
    earnings, but also increased social capital, which would provide
    tenants with additional supports if facing eviction.

4.  **Income Level**

    Incomes are expected to correlate with eviction filings, as lower
    income residents are more likely to be at risk of evictions due to
    non-payment of rent.

5.  **Age of Housing Stock**

    Housing stock age is included under the rationalization that newer,
    higher quality, units are more likely to evict tenants for lease
    infringements or damage to the property. Furthermore, new units are
    expected to have stronger / professionalized property management
    teams capable of filing for eviction more efficiently.

6.  **Proportion of Renters**

    Census tracts with a large share of renters are more likely to have
    higher eviction rates, as homeowners cannot be evicted.

7.  **Unemployment Rate**

    Similar to income, we would expect unemployed renters to be unable
    to cover portions of their rent, thus resulting in late payments,
    arrears, and eventual eviction.

8.  **Racial Majority**

    We want to include a racial analysis, as Black, Hispanic, and other
    minority groups are likely at higher risk for eviction due to
    systemic inequality.

9.  **Gender of Evictions**

    Similar to race, we want to see if gender is a correlate with
    evictions. This variable represents the proportion of the census
    tract that are males.

10. **Temporal Lags**

    Given that rental markets operate cyclically, we included a monthly
    and yearly lag to account for this.

11. **Housing Choice Voucher**

    We included HCV recipients by census tract to attempt to complicate
    the correlation between low-incomes and evictions. We assume that
    HCV recipients, though low income, are less likely to be evicted, as
    they are not expected to pay more than 30% of their income in rent.
    This helps to control for shocks experienced to households such as
    job loss or medical emergencies.

12. **Proportion of Vacant Lots**

    Our inclusion of vacant land rests on the assumption that fewer
    vacancies indicates a more competitive or built-out rental market,
    and thus landlords are more likely to advance eviction proceedings,
    as they are more likely to attract a replacement tenant.
    Furthermore, high lot vacancy rates may be indicative of absentee
    landlords who are less likely to file for eviction.

13. **Month Fixed Effects**

    Finally, month fixed effects controls for factors outside of these
    variables which occur monthly. For example, January sees high levels
    of evictions, likely due to court processing timelines or leases
    expiring.

```{r}
#| code-fold: true
#| code-summary: "Gathering Census Data"
#| message: false
# census data

# Load variables
acs_vars_2024 <- load_variables(2024, "acs5", cache = TRUE)
#acs_vars_2020 <- load_variables(2020, "acs5", cache = TRUE)

# Load secondary data on access + socioeconomics + amenities + spacial structure
# Define ACS variables 2024
census_vars <- c(
  med_inc      = "B19013_001",   # Median household income
  med_home_value  = "B25077_001",   # Median home value
  
  total_pop       = "B01003_001",   # Total population
  poverty_total   = "B17001_002",   # Poverty universe
  
  edu_bachelors   = "B15003_022",   # Bachelor's
  edu_masters     = "B15003_023",
  edu_prof        = "B15003_024",
  edu_phd         = "B15003_025",
  
  tenure_total    = "B25003_001",   # Total occupied
  renter_occupied  = "B25003_003",    # renter occupied
  
  white = "B03002_003",
  black = "B03002_004",
  latinx = "B03002_012",
  
  rent_3034 = "B25070_007", # rent burden
  rent_3539 = "B25070_008",
  rent_4049 = "B25070_009",
  rent_50 = "B25070_010",
  
  med_rent = "B25064_001",
  
  house_age = "B25035_001", # median housing built year
  
  unemp = "B23025_005",

  pop_male = "B01001_002"
)

# Pull census data
phl_census <- get_acs(
  geography = "tract",
  variables = census_vars,
  state = "PA",
  county = "Philadelphia",
  year = 2024,
  survey = "acs5",
  geometry = TRUE,
  progress = FALSE,
  
)

phl_census_clean <- phl_census %>% 
  select(GEOID, NAME, variable, estimate, geometry) %>% # Drop MOE
  pivot_wider(names_from = variable, values_from = estimate) %>% # Pivot to wide table
  mutate(
    prop_male = pop_male/total_pop,
    edu_higher = (edu_bachelors + edu_prof + edu_phd)/total_pop,
    rent_burden = (rent_3034 + rent_3539 + rent_4049 + rent_50)/total_pop,
    pov_rate = poverty_total/total_pop,
    rent_occ = renter_occupied/tenure_total,
    white = white/total_pop,
    black = black/total_pop,
    latinx = latinx/total_pop,
    unemp = unemp/total_pop,
    house_age = 2026-house_age
  ) %>%
  select(
    GEOID, NAME, geometry, total_pop, prop_male, white, black, latinx, edu_higher, rent_burden, med_rent, pov_rate, med_inc, 
    unemp, rent_occ, house_age, med_home_value
  ) %>%
  st_transform(4326)
  
hcv <- st_read(here("data/philly_hcv_data.shp"), quiet = TRUE) %>% filter(!grepl("34", STATE)) %>% st_transform(4326)
vacant <- st_read(here("data/Vacant_Indicators_Bldg.shp"), quiet = TRUE) %>% st_transform(4326)

#Grouping vacancies by census tract
vacant_spatial <- vacant %>%
  st_join(tracts_phl)

vacant_spatial <- vacant_spatial %>%
  st_drop_geometry() %>%
  group_by(GEOID) %>%
  summarize(n_vacant = n())

hcv_vacant <- hcv %>%
  left_join(vacant_spatial, by = "GEOID") %>%
  select(GEOID, HCV_PUBL_1, HCV_PUBLIC, n_vacant)

hcv_vacant <- hcv_vacant %>%
  mutate(n_vacant = replace_na(n_vacant, 0),
         HCV_PUBL_1 = (HCV_PUBL_1 / 100))

# join hcv and vacant lots onto census data

predictor_varbs <- phl_census_clean %>%
  st_drop_geometry() %>%
  left_join(hcv_vacant, by = "GEOID")

predictor_varbs <- predictor_varbs %>%
  drop_na()

#Now we have a final dataset with all our predictor variables that can be joined to the evictions data
#Joining with our evictions dataset for final model building. 

eviction <- eviction %>%
  left_join(predictor_varbs, by = "GEOID")

eviction <- eviction %>%
  st_as_sf()
```

## Spatial Patterns of Census Data

We assumed that vulnerable populations are more likely to face
evictions. Due to this, we examined the proportion of Black population,
those living rent burdened, and those living below the poverty line of
each census. These spatial distributions were then compared against the
distribution of evictions throughout Philadelphia. Visualizing the
distribution of Philadelphia's Black population, this revealed
significant clustering in North and West Philadelphia - areas that
should be examined closer for elevated eviction filings.

```{r}
#| code-fold: true
#| code-summary: "Exploring Census data"
#| message: false
# demographic characteristics vs eviction filings
demo_map <- ggplot(phl_census_clean) +
  geom_sf(aes(fill = black), color = NA) +
  scale_fill_gradient2(low = "black",
                       high = "blue",
                       na.value = "white",
                       name = "Proportion of Black Population") +
  labs(title = "Spatial Distribution of Racial Demographics")+
  theme_minimal()

rent_burden_map <- ggplot(phl_census_clean) +
  geom_sf(aes(fill = rent_burden), color = NA) +
  scale_fill_gradient2(low = "black",
                       high = "purple",
                       na.value = "white",
                       name = "Proportion Rent Burdened") +
  labs(title = "Spatial Distribution of Rent Burdened Population")+
  theme_minimal()

poverty_rate <- ggplot(phl_census_clean) +
  geom_sf(aes(fill = pov_rate), color = NA) +
  scale_fill_gradient2(low = "black",
                       high = "darkgreen",
                       na.value = "white",
                       name = "Proportion Below Poverty Line") +
  labs(title = "Spatial Distribution of Population Living Below Poverty Line")+
  theme_minimal()

demo_map
```

Rent burden reveals a different spatial pattern. While majority Black
census tracts have a large proportion being rent burdened, especially in
West Philadelphia, the highest levels of rent burden are in expensive
neighborhoods such as Rittenhouse, Fitler Square, Washington Square
West, and Farimount. This may reflect the fact that these neighborhoods
have overpriced baseline housing costs.

```{r rent burden}
#| code-summary: "Exploring Census data"
#| message: false
rent_burden_map
```

With the exception of West Philadelphia, cost burden and poverty rate do
not appear to be strongly spatially correlated. However, majority black
census tracts appear much more likely to have a large share of residents
living below the poverty line.

```{r}
#| code-summary: "Exploring Census data"
#| message: false
poverty_rate
```

Given our exploration of census variables that we believe are strong
predictors of evictions, we expect that West Philadelphia is at the
greatest risk of elevated eviction filings due to its simultaneous
clustering of majority black census tracts, cost burden, and poverty
rates.

## Spatial Patterns in Evictions

```{r exploring evictions}
#| code-summary: "Exploring Eviction data"
#| message: false

# filter to 2025 only
evictions_2025 <- spatial_evictions %>%
  filter(year == 2025)

evictions_2025_proj <- st_transform(evictions_2025, 26918)  # NAD83 / UTM zone 18N

evictions_dots <- evictions_2025_proj %>%
  mutate(n_dots = filings_2020) %>%
  filter(n_dots > 0) %>%
  mutate(points = map2(geometry, n_dots, ~ st_sample(.x, size = .y))) %>%
  st_drop_geometry() %>%
  unnest(points) %>%
  st_as_sf() %>%
  st_set_crs(26918)  # WGS84 (lat/long)

# plot
evictions_map <- ggplot() +
  geom_sf(data = evictions_2025_proj, fill = "grey95", color = "white") +
  geom_sf(data = evictions_dots, color = "red", size = 0.2, alpha = 0.6) +
  theme_minimal() +
  labs(
    title = "Eviction Filings (2025)",
    subtitle = "Dot density map"
  )

evictions_and_black_populatoin <- ggplot(phl_census_clean) +
  geom_sf(aes(fill = black), color = NA) +
  geom_sf(data = evictions_dots, color = "red", size = 0.002, alpha = 0.5) +
  scale_fill_gradient2(low = "black",
                       high = "blue",
                       na.value = "white",
                       name = "Proportion of Black Population") +
  labs(title = "Spatial Distribution of Racial Demographics vs Eviction Filings")+
  theme_minimal()

evictions_and_rent_burden <- ggplot(phl_census_clean) +
  geom_sf(aes(fill = rent_burden), color = NA) +
  geom_sf(data = evictions_dots, color = "red", size = 0.002, alpha = 0.5) +
  scale_fill_gradient2(low = "black",
                       high = "purple",
                       na.value = "white",
                       name = "Proportion Rent Burdened") +
  labs(title = "Spatial Distribution of Rent Burdened Population vs Eviction Filings")+
  theme_minimal()

poverty_evictions <- ggplot(phl_census_clean) +
  geom_sf(aes(fill = pov_rate), color = NA) +
  geom_sf(data = evictions_dots, color = "red", size = 0.002, alpha = 0.5) +
  scale_fill_gradient2(low = "black",
                       high = "darkgreen",
                       na.value = "white",
                       name = "Proportion Below Poverty Line") +
  labs(title = "Spatial Distribution of Population Living Below Poverty Line vs Eviction Filings")+
  theme_minimal()

```

```{r poverty rate}
#| code-summary: "Eviction data"
#| message: false
evictions_map
```

Eviction filings are spatially dispersed through residential areas
throughout the city. However, on first inspection it appears that there
are significantly more evictions filed in North and West Philadelphia,
consistent with our census data hypothesis.

```{r}
#| code-summary: "Eviction data"
#| message: false
evictions_and_black_populatoin
```

Examining the spatial distribution of evictions and majority black
census tracts, there is a noticeable overlap between majority black
census tracts and the prevalence of evictions. However, large portions
such as Kensington and the Riverwards, as well as areas of South
Philadelphia and East Falls, do not have a sizable black population but
appear to have elevated eviction filings.

```{r}
#| code-summary: "Eviction data"
#| message: false
evictions_and_rent_burden
```

Examining evictions and rent burden shows that there are key parts of
the city with low eviction levels and high degrees of rent burden. In
particular, portions of the Northeast, Center City, and Queen Village
have high levels of rent burden but low eviction counts.

```{r}
#| code-summary: "Eviction data"
#| message: false
poverty_evictions
```

Evictions filings and the proportion of residents below the poverty line
appear to be fairly correlated. A noteworthy exception to this is a
portion of Southwest Philadelphia that has high levels of poverty but
low eviction filings.

```{r}
#| code-fold: true
#| code-summary: "Developing Complete Panel Dataset"
#| message: false
#Now I need to create the panel dataset in order to build the model and do train/test splits

#Data is currently in a station-month structure, where each row represents a unique station-month combination. 

# Calculate expected panel size
n_months <- eviction %>%
  distinct(month, year) %>%
  nrow()
n_tracts <- length(unique(eviction$GEOID))
expected_rows <- n_months * n_tracts

#We expect the eviction dataset to have 31,008 rows based on the total number of possible month / tract combinations. Given that the dataset already has this number of rows, we can conclude that it is already a complete panel data set, and will not require expanding the panel to include missing combinations. 

#However, we do want to cut the evictions data down to remove outliers that we observed in our EDA. Specifically, we only want ot October 2022 and forward, as this is seemingly when the evictions stabilized. 

eviction_final <- eviction %>%
  filter(year > 2021 | (year == 2021 & month >= 10))

#making a temporal lag for one month ago and one year ago

# Create lag variables for each tract
eviction_final <- eviction_final %>%
  arrange(GEOID, year, month) %>%
  group_by(GEOID) %>%
  mutate(
    lag1_month = lag(filings_2020, 1),   # evictions 1 month ago
    lag12_year = lag(filings_2020, 12)   # evictions 1 year ago (same month)
  ) %>%
  ungroup()

# Remove rows with NA lags (first 24 hours for each station)
eviction_final <- eviction_final %>%
  filter(!is.na(lag1_month), !is.na(lag12_year))

#need to remove some census tracts from the data that had missing census data or vacancies data. 
eviction_final <- eviction_final %>%
  filter(!is.na(total_pop))
```

```{r train-test split}
#| code-fold: true
#| code-summary: "train-test split"
#| message: false
# Which stations have trips in BOTH early and late periods?
early_evictions <- eviction_final %>%
  filter(year < 2024) %>%
  filter(filings_2020 > 0) %>%
  distinct(GEOID) %>%
  pull(GEOID)

late_evictions <- eviction_final %>%
  filter(year >= 2024) %>%
  filter(filings_2020 > 0) %>%
  distinct(GEOID) %>%
  pull(GEOID)

# Keep only stations that appear in BOTH periods
common_evictions <- intersect(early_evictions, late_evictions)


# Filter panel to only common stations
eviction_model_data <- eviction_final %>%
  filter(GEOID %in% common_evictions)

# NOW create train/test split
train <- eviction_model_data %>%
  filter(year < 2024)

test <- eviction_model_data %>%
  filter(year >= 2024)
```

# Model Building

## Temporal Lag Model

Our first model includes only temporal lags as predictors of future
evictions. We began by using a poisson model, but analysis revealed that
the data is over-distributed, thus we switched to using a negative
binomial model.

```{r temporal model}
#| code-fold: true
#| code-summary: "temporal lag"
#| message: false
#now building the first model that only uses temporal lag variables

temporal_only_model <- glm(
  filings_2020 ~ lag1_month + lag12_year,
  data = train,
  family = poisson
)

temporal_only_model

#given that residual deviance is much greater than the degrees of freedom, we should use a negative binomial model for this data
library(MASS)

temporal_only_model_nb <- glm.nb(filings_2020 ~ lag1_month + lag12_year,
                                 data = train)

summary(temporal_only_model_nb)
#the negative binomial model's residual deviance is much closer to the degrees of freedom, making this a stronger model
```

## Census Model

Our next model builds on the temporal lag model, but includes census
variables such as: racial demographics, higher education attainment,
poverty rate, rent burden, income levels, unemployment rate, housing
age, home values, and gender ratios.

Importantly this model finds that educational attainment, cost burden,
poverty rate, and race are significant predictors of eviction.

```{r census model}
#| code-fold: true
#| code-summary: "census model"
#| message: false
census_model <- glm.nb(filings_2020 ~ lag1_month + lag12_year + white + edu_higher + rent_burden + pov_rate + med_inc + unemp + house_age + med_home_value + prop_male,
                                 data = train)

summary(census_model)
```

## Expanded Census and Neighborhood Factors Model

The final model includes all of the variables outlined above along with
housing choice vouchers, vacancies, and monthly fixed effects.

```{r everything model}
#| code-fold: true
#| code-summary: "expanded census and neighborhood factors model"
#| message: false
everything_model <- glm.nb(filings_2020 ~ lag1_month + lag12_year + white + edu_higher + rent_burden + pov_rate + med_inc + unemp + house_age + med_home_value + prop_male + n_vacant + HCV_PUBL_1 + as.factor(month),
                                 data = train)

summary(everything_model)
```

# Initial Model Comparisons

Our initial models saw modest improvements as new variables were added.
However, with an MAE of approximately 2 evictions, while mean evictions
by census tracts is also approximately 2, these models have very little
to no predictive capacity. Thus, we pursued binary predictions to more
accurately classify tracts according to risk.

```{r MAEs}
#| code-fold: true
#| code-summary: "initial model comparisons"
#| message: false

test <- test %>%
  mutate(
    pred1 = predict(temporal_only_model, newdata = test),
    pred2 = predict(temporal_only_model_nb, newdata = test),
    pred3 = predict(census_model, newdata = test),
    pred4 = predict(everything_model, newdata = test)
  )

# Calculate MAE for each model
mae_results <- data.frame(
  Model = c(
    "1. Poisson Model - Temporal Lag Only",
    "2. Negative Binomial - Temporal Lag Only",
    "3. Negative Binomial - Temporal Lag + Census",
    "4. Negative Binomial - Temporal Lag + Census + Other"
    
  ),
  MAE = c(
    mean(abs(test$filings_2020 - test$pred1), na.rm = TRUE),
    mean(abs(test$filings_2020 - test$pred2), na.rm = TRUE),
    mean(abs(test$filings_2020 - test$pred3), na.rm = TRUE),
    mean(abs(test$filings_2020 - test$pred4), na.rm = TRUE)
  )
)

kable(mae_results, 
      digits = 2,
      caption = "Mean Absolute Error by Model (Test Set)",
      col.names = c("Model", "MAE (evictions)")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Where are Model Errors Most Concentrated?

Based on the map below, errors are most concentrated in Northeast,
Northwest, and West Philadelphia. This indicates that our model is
failing to capture local variation in these parts of the city.

```{r visualizing MAE}
#| code-fold: true
#| code-summary: "Visualizing MAE"
#| message: false
tract_errors <- test %>%
  group_by(GEOID) %>%
  summarize(
    MAE = mean(abs(filings_2020 - pred1), na.rm = TRUE),
    avg_demand = mean(filings_2020, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(!is.na(GEOID))
error_plot <- ggplot() +
  geom_sf(data = tracts_phl, color = "gray80", fill = "gray95", linewidth = 0.2) +
  geom_sf(data = tracts_phl %>% left_join(st_drop_geometry(tract_errors), by = "GEOID"),
          aes(fill = MAE), color = "white", linewidth = 0.2) +
  scale_fill_viridis_c(
    option = "magma",
    direction = -1,
    na.value = "gray85",
    name = "MAE"
  ) +
  labs(
    title = "Model Error by Census Tract",
    subtitle = "Mean Absolute Error in predicted eviction filings",
    caption = "Gray tracts had no predictions"
  ) +
  theme_void(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5, color = "gray40"),
    legend.position = "right"
  )

error_plot
```

# Logistic Model

Given that the poisson and negative binomial models are not strong
predictors of evictions by census tract, we developed a logistic model
to predict the census tracts that are predicted to be at high risk of
eviction filings.

In doing so, we had to define what was classified as "high risk." Census
tracts that are 1.5 standard deviations above the average are considered
high-eviction tracts for the purposes of our classification. We
considered these tracts to be anomalous among census tracts, and thus
need targeted intervention.

```{r plotting evictions data}
#| code-fold: true
#| code-summary: "Evictions Panel Distribution"
#| message: false
#fist finding tracts that are in the top quartile by month 
ggplot(eviction_model_data, aes(x = filings_2020)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "white") +
  labs(
    title = "Distribution of Eviction Filings (2020)",
    x = "Filings (2020)",
    y = "Count"
  ) +
  theme_minimal()

```

## Model Predictive Capacity and Limitations

Our logistic model has strong predictive capacity, with a AUC of 0.898.
This means that our model is able to accurately classify at a much
higher rate than random chance. Our model also strongly predicts census
tracts with a low eviction counts, but this is largely due to our data
skewing towards low-risk tracts. However, our model only accurately
predicts roughly 60% of true positives, indicating that the model should
be refined further to better predict high-risk census tracts.

We chose a threshold of 0.75 for predicting low risk census tracts,
which means that we were willing to risk more false positives at the
expense of lower negative accuracy. This is because we'd rather falsely
predict a high-risk census tract than a low-risk census tract.

```{r}
#| code-fold: true
#| code-summary: "Lodistic Model"
#| message: false

eviction_model_data <- eviction_model_data %>%
  group_by(month) %>%
  mutate(category = as.factor(ifelse(filings_2020 > mean(filings_2020, na.rm = TRUE) + 1.5 * sd(filings_2020, na.rm = TRUE), "high", "low"))) %>%
  ungroup()

set.seed(123)
split <- initial_split(eviction_model_data, prop = 0.05, strata = category)
train2 <- training(split)
test2  <- testing(split)

train2 <- train2 %>% mutate(month  = as.factor(month))
test2 <- test2 %>% mutate(month = as.factor(month))

#using the tidymodels package

#creating the model
logit_recipe <- recipe(category ~ lag1_month + lag12_year + white + edu_higher + rent_burden + pov_rate + med_inc + unemp + house_age + med_home_value + prop_male + n_vacant + HCV_PUBL_1 + month, data = train2) %>%
  step_normalize(all_numeric_predictors()) %>%
  step_dummy(all_nominal_predictors())


logit_spec <- logistic_reg() %>%
  set_engine("glm") %>%
  set_mode("classification")

logit_wf <- workflow() %>%
  add_recipe(logit_recipe) %>%
  add_model(logit_spec)

logit_fit <- logit_wf %>% fit(data = train2)

test2 <- test2 %>%
  bind_cols(predict(logit_fit, test2, type = "prob")) %>%
  bind_cols(predict(logit_fit, test2))

test2_df <- test2 %>% sf::st_drop_geometry()

conf_mat(test2_df, truth = category, estimate = .pred_class)
roc_auc(test2_df, truth = category, .pred_high)

#Creating a lower threshold so that we capture more true positives at the risk of also capturing more false positives
library(yardstick)

test2_df <- test2_df %>%
  mutate(.pred_class_low_threshold = as.factor(ifelse(.pred_high >= 0.75, "high", "low")))

cm <- conf_mat(test2_df, truth = category, estimate = .pred_class_low_threshold)

confusion_matrix <- autoplot(cm, type = "heatmap") +
  scale_fill_gradient(low = "#E6F1FB", high = "#185FA5") +
  labs(
    title = "Confusion Matrix",
    subtitle = "Logistic model — high vs. low eviction filings",
    x = "Predicted",
    y = "Actual"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5, color = "gray50"),
    legend.position = "none",
    panel.grid = element_blank()
  )

# --- ROC Curve ---
auc_val <- roc_auc(test2_df, truth = category, .pred_high) %>% pull(.estimate) %>% round(3)


ROC_Curve <- roc_curve(test2_df, truth = category, .pred_high) %>%
  ggplot(aes(x = 1 - specificity, y = sensitivity)) +
  geom_line(color = "#185FA5", linewidth = 1.2) +
  geom_abline(linetype = "dashed", color = "gray70") +
  annotate("text", x = 0.75, y = 0.1,
         label = paste0("AUC = ", auc_val),
           size = 4.5, color = "#185FA5", fontface = "bold") +
  labs(
    title = "ROC Curve",
    subtitle = "Logistic model — high vs. low eviction filings",
    x = "False positive rate (1 - specificity)",
    y = "True positive rate (sensitivity)"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5, color = "gray50"),
    panel.grid.minor = element_blank()
  )

confusion_matrix | ROC_Curve
```

# Policy Implications and Recommended Implementation

Our model demonstrates that tracts with high eviction counts can be
predicted. Though our model is limited in the high-risk tracts that it
captures, it nonetheless captures 60% of high-risk tracts, meaning that
interventions can be targeted at a majority of high-risk tracts. Our
model's ability to strongly predict low-risk tracts is not a drawback,
however, as this ensures that the city would not be over-extending
resources to areas of the city that may not need it as much.

Furthermore, our exploratory analysis finds that rent burden is a strong
predictor of eviction likelihood, meaning that policies which strengthen
incomes or reduce rent will positively impact evictions city wide. There
were also several tracts where rent burden was high, but evictions were
low, implying some different relationship at play mediating rent burden
and evictions, and this is an avenue for further explorations. Either by
implementing policies as mentioned, or examining why certain
rent-burdened tracts are more likely to evict their tenants, it is
necessary to understand this distinction, so as to ensure even the
rent-burdened are housed, especially in high-risk tracts.

The uneven spread of our model error across the city is also a point of
concern and is something to be explored. The areas of Northeast,
Northwest, and West Philadelphia are where the errors were most heavily
concentrated, and this is concerning especially as these are the areas
where we determined the vulnerable population to be concentrated, and
where there are higher rates of evictions. High errors make for
unreliable models, and in cases where people's livelihoods are
concerned, this is an alarming point that needs to be addressed. The
unequal distribution of errors implies an underlying problem either
during the data collection or data processing phases. The fact that this
is an issue, especially in vulnerable areas, is an inequity issue that
needs to be remediated.

Therefore, we recommend that the data collection and processing stages
are re-evaluated to ensure equitable data collection throughout the
entire city to best recreate our model, which in turn will yield more
tangible results, informing nuanced policy decisions.

We also recommend that the model is updated monthly as eviction reports
are released by the Princeton Eviction Lab, allowing the model to
provide a monthly prediction of which census tracts in the city are at
particular risk of eviction.

# Data Sources

Eviction Data - Eviction Lab, Philadelphia, Monthly Data -
evictionlab.org

Census Data - Gender (Sex), Household Tenure (Renters vs. Owners), Race,
Unemployment, Housing Age (Current Year - Year Built), Educational
Attainment, Housing Costs, Rent Burden - ACS, 2024, Census Tracts

OpenDataPhilly - Vacancy - opendataphilly.org

Housing Choice Voucher - HCV Data - hud.gov
