---
title: "Lab 4: Spatial Predictive Analysis"
subtitle: "Analyzing 311 Violations"
author: "Angie Kwon"
date: today
format: 
  html:
    code-fold: true
    toc: true
    toc-location: left
    theme: cosmo
    embed-resources: true
execute:
  warning: false
  message: false
---

## Assignment Overview

In this lab, you will apply the spatial predictive modeling techniques demonstrated in the class exercise using a different 311 service request type of your choice as the predictor variable. You will build a complete spatial predictive model, document your process, and interpret your results.

### Learning Objectives

By completing this assignment, you will demonstrate your ability to:

-   Adapt example code to analyze a new dataset
-   Build spatial features for predictive modeling
-   Apply count regression techniques to spatial data
-   Implement spatial cross-validation
-   Interpret and communicate model results
-   Critically evaluate model performance

------------------------------------------------------------------------

```{r}
#| include: false

# Load required packages
library(tidyverse)      # Data manipulation
library(sf)             # Spatial operations
library(here)           # Relative file paths
library(viridis)        # Color scales
library(terra)          # Raster operations (replaces 'raster')
library(spdep)          # Spatial dependence
library(FNN)            # Fast nearest neighbors
library(MASS)           # Negative binomial regression
library(patchwork)      # Plot composition (replaces grid/gridExtra)
library(knitr)          # Tables
library(kableExtra)     # Table formatting
library(classInt)       # Classification intervals


# Spatstat for analyzing point patterns - split into sub-packages
library(spatstat.geom)    # Spatial geometries
library(spatstat.explore) # Spatial exploration/KDE

setwd("/Users/angiekwon/Documents/GitHub/akwon_portfolio/labs/lab_4")
#setwd("C:/Users/angel/Documents/akwon_portfolio/labs/lab_4")

theme_crime <- function(base_size = 11) {
  theme_minimal(base_size = base_size) +
    theme(
      plot.title = element_text(face = "bold", size = base_size + 1),
      plot.subtitle = element_text(color = "gray30", size = base_size - 1),
      legend.position = "right",
      panel.grid.minor = element_blank(),
      axis.text = element_blank(),
      axis.title = element_blank()
    )
}

# Set as default
theme_set(theme_crime())
```

## The Analysis

Using the class exercise as your template, adapt the code to analyze your chosen 311 violation type.

#### Part 1: Data Loading & Exploration

-   Load your 311 data and Chicago spatial boundaries
-   Create visualizations showing the spatial distribution of your violation type
-   Describe patterns you observe

I chose alley light outage as a predictor for burglaries in Chicago, because I felt that the broken window effect would be most present in a place where the disarray was most salient, especially during night which is when more crime occurs. Alleys are already untrustworthy as public spaces and broken lights would only add to the feel of unsafety.

```{r}
#| include: false

lights_locations <- read_csv("data/311_Service_Requests_-_Alley_Lights_Out_-_Historical_20260408.csv") %>%
  rename_with(~ gsub(" ", "_", .x)) %>%
  filter(!is.na(Latitude), !is.na(Longitude)) %>%
  st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326) %>%
  st_transform('ESRI:102271')
  
# Load Chicago boundary
chicagoBoundary <- 
  st_read("https://raw.githubusercontent.com/urbanSpatial/Public-Policy-Analytics-Landing/master/DATA/Chapter5/chicagoBoundary.geojson") %>%
  st_transform('ESRI:102271')

# burglaries
burglaries <- st_read(here("inclass_exercise/week9/data/burglaries.shp")) %>% 
  st_transform('ESRI:102271')

# police districts (used for spatial cross-validation)
policeDistricts <- 
  st_read("https://data.cityofchicago.org/api/geospatial/24zt-jpfn?method=export&format=GeoJSON") %>%
  st_transform('ESRI:102271') %>%
  dplyr::select(District = dist_num)

# vacant buildings
vacant <- st_read("data/geo_export_656aaaf5-307a-4116-9ebc-a5d8299babd8.shp") %>%
  st_transform('ESRI:102271')
```

```{r}
#| echo: false

# Check the data
cat("\n✓ Loaded alley lights-out data\n")
cat("  - Number of lights out:", nrow(lights_locations), "\n")
cat("  - CRS:", st_crs(lights_locations)$input, "\n")
cat("  - Date range:", format(min(lights_locations$Creation_Date, na.rm = TRUE)), "to", 
    format(max(lights_locations$Creation_Date, na.rm = TRUE)), "\n")

# plot points
p1 <- ggplot() + 
  geom_sf(data = chicagoBoundary, fill = "gray95", color = "gray60") +
  geom_sf(data = lights_locations, color = "#d62828", size = 0.1, alpha = 0.4) +
  labs(
    title = "Alley Lights-Out Locations",
    subtitle = paste0("Chicago 2017, n = ", nrow(lights_locations))
  )

# density surface using modern syntax
p2 <- ggplot() + 
  geom_sf(data = chicagoBoundary, fill = "gray95", color = "gray60") +
  geom_density_2d_filled(
    data = data.frame(st_coordinates(lights_locations)),
    aes(X, Y),
    alpha = 0.7,
    bins = 8
  ) +
  scale_fill_viridis_d(
    option = "plasma",
    direction = -1,
    guide = "none"  # Modern ggplot2 syntax (not guide = FALSE)
  ) +
  labs(
    title = "Density Surface",
    subtitle = "Kernel density estimation"
  )

# Combine plots using patchwork (modern approach)
p1 + p2 + 
  plot_annotation(
    title = "Spatial Distribution of Alley Light-Outs in Chicago",
    tag_levels = 'A'
  )
```

There seems to be clear hotspots, especially in South Chicago, with a central coldspot on the central East side.

#### Part 2: Fishnet Grid Creation

-   Create a 500m x 500m fishnet grid
-   Aggregate your violations to grid cells
-   Visualize the count distribution

```{r create-fishnet}
# Create 500m x 500m grid
fishnet <- st_make_grid(
  chicagoBoundary,
  cellsize = 500,  # 500 meters per cell
  square = TRUE
) %>%
  st_sf() %>%
  mutate(uniqueID = row_number()) %>%
  st_transform('ESRI:102271') 

# Keep only cells that intersect Chicago
fishnet <- fishnet[chicagoBoundary, ]

# View basic info
cat("✓ Created fishnet grid\n")
cat("  - Number of cells:", nrow(fishnet), "\n")
cat("  - Cell size:", 500, "x", 500, "meters\n")
cat("  - Cell area:", round(st_area(fishnet[1,])), "square meters\n")
```

```{r aggregate-lights-out}
# Spatial join: which cell contains each burglary?
lights_fishnet <- st_join(lights_locations, fishnet, join = st_within) %>%
  st_drop_geometry() %>%
  group_by(uniqueID) %>%
  summarize(countLights = n()) 

glimpse(lights_fishnet)
st_crs(lights_fishnet)

# Join back to fishnet (cells with 0 burglaries will be NA)
fishnet <- fishnet %>%
  left_join(lights_fishnet, by = "uniqueID") %>%
  mutate(countLights = replace_na(countLights, 0))

# Summary statistics
cat("\nLight-out count distribution:\n")
summary(fishnet$countLights)
cat("\nCells with zero burglaries:", 
    sum(fishnet$countLights == 0), 
    "/", nrow(fishnet),
    "(", round(100 * sum(fishnet$countLights == 0) / nrow(fishnet), 1), "%)\n")
```

```{r aggregate-burglaries}
# Spatial join: which cell contains each burglary?
burglaries_fishnet <- st_join(burglaries, fishnet, join = st_within) %>%
  st_drop_geometry() %>%
  group_by(uniqueID) %>%
  summarize(countBurglaries = n()) 


glimpse(burglaries_fishnet)
st_crs(burglaries_fishnet)

# Join back to fishnet (cells with 0 burglaries will be NA)
fishnet <- fishnet %>%
  left_join(burglaries_fishnet, by = "uniqueID") %>%
  mutate(countBurglaries = replace_na(countBurglaries, 0))

# Summary statistics
cat("\nBurglary count distribution:\n")
summary(fishnet$countBurglaries)
cat("\nCells with zero burglaries:", 
    sum(fishnet$countBurglaries == 0), 
    "/", nrow(fishnet),
    "(", round(100 * sum(fishnet$countBurglaries == 0) / nrow(fishnet), 1), "%)\n")
```

```{r kde-baseline}
#| message: false

# Convert burglaries to ppp (point pattern) format for spatstat
burglaries_ppp <- as.ppp(
  st_coordinates(burglaries),
  W = as.owin(st_bbox(chicagoBoundary))
)

# Calculate KDE with 1km bandwidth
kde_burglaries <- density.ppp(
  burglaries_ppp,
  sigma = 1000,  # 1km bandwidth
  edge = TRUE    # Edge correction
)

# Convert to terra raster (modern approach, not raster::raster)
kde_raster <- rast(kde_burglaries)

# Extract KDE values to fishnet cells
fishnet <- fishnet %>%
  mutate(
    kde_value = terra::extract(
      kde_raster,
      vect(fishnet),
      fun = mean,
      na.rm = TRUE
    )[, 2]  # Extract just the values column
  )

cat("✓ Calculated KDE baseline\n")
```

```{r visualize-kde}
#| fig-width: 8
#| fig-height: 6

ggplot() +
  geom_sf(data = fishnet, aes(fill = kde_value), color = NA) +
  #geom_sf(data = chicagoBoundary, fill = "blue", color = "white", linewidth = 1) +
  scale_fill_viridis_c(
    name = "KDE Value",
    option = "plasma"
  ) +
  labs(
    title = "Kernel Density Estimation Baseline",
    subtitle = "Simple spatial smoothing of burglary locations"
  ) +
  theme_crime()
```

```{r aggregate-vacant-buildings}
vacant_fishnet <- st_join(vacant, fishnet, join = st_within) %>%
  st_drop_geometry() %>%
  group_by(uniqueID) %>%
  summarize(countVacant = n()) 


glimpse(vacant_fishnet)
st_crs(vacant_fishnet)

# Join back to fishnet (cells with 0 burglaries will be NA)
fishnet <- fishnet %>%
  left_join(vacant_fishnet, by = "uniqueID") %>%
  mutate(countVacant = replace_na(countVacant, 0))

# Summary statistics
cat("\nVacant count distribution:\n")
summary(fishnet$countVacant)
cat("\nCells with zero vacant buildings:", 
    sum(fishnet$countVacant == 0), 
    "/", nrow(fishnet),
    "(", round(100 * sum(fishnet$countVacant == 0) / nrow(fishnet), 1), "%)\n")
```

```{r visualize-fishnet}
#| fig-width: 8
#| fig-height: 6

# Visualize aggregated counts
ggplot() +
  geom_sf(data = fishnet, aes(fill = countLights), color = NA) +
  geom_sf(data = chicagoBoundary, fill = NA, color = "white", linewidth = 1) +
  scale_fill_viridis_c(
    name = "Alley Light-Out",
    option = "plasma",
    trans = "sqrt",  # Square root for better visualization of skewed data
    breaks = c(0, 25, 50, 100, 200, 300, 400, 500, 600)
  ) +
  labs(
    title = "Alley Light-Outs Counts by Grid Cell",
    subtitle = "500m x 500m cells, Chicago 2017"
  ) +
  theme_crime()
```

Creating a fishnet allows us to analyze data where they are spatially in regular cells so that the data isn't weighted by the different sizes people generally use for boundaries on maps. This also allows us to have concrete data for each cells, rather than a general picture like we get in heat maps.

#### Part 3: Spatial Features

-   Calculate k-nearest neighbor features
-   Perform Local Moran's I analysis
-   Identify hot spots and cold spots
-   Create distance-to-hotspot measures
-   Join any additional contextual data if you are looking for more to do and really get into this (e.g., demographics, land use)

```{r nn-feature}
#| message: false

# Calculate mean distance to 3 nearest abandoned cars
# (Do this OUTSIDE of mutate to avoid sf conflicts)

# Get coordinates
fishnet_coords <- st_coordinates(st_centroid(fishnet))
lights_coords <- st_coordinates(lights_locations)

# Calculate k nearest neighbors and distances
nn_result <- get.knnx(lights_coords, fishnet_coords, k = 3)

# Add to fishnet
fishnet <- fishnet %>%
  mutate(
    lights_locations.nn = rowMeans(nn_result$nn.dist)
  )

cat("✓ Calculated nearest neighbor distances\n")
summary(fishnet$lights_locations.nn)
```

Calculating the 3 nearest broken alley lights from each fishnet cell's center. Then, calculating the mean distance to each of them to gauge how close all 3 are to the center, and therefore how common and close together they are.

```{r local-morans-lights}
# Function to calculate Local Moran's I
calculate_local_morans <- function(data, variable, k = 5) {
  
  # Create spatial weights
  coords <- st_coordinates(st_centroid(data))
  neighbors <- knn2nb(knearneigh(coords, k = k))
  weights <- nb2listw(neighbors, style = "W", zero.policy = TRUE)
  
  # Calculate Local Moran's I
  local_moran <- localmoran(data[[variable]], weights)
  
  # Classify clusters
  mean_val <- mean(data[[variable]], na.rm = TRUE)
  
  data %>%
    mutate(
      local_i = local_moran[, 1],
      p_value = local_moran[, 5],
      is_significant = p_value < 0.05,
      
      moran_class = case_when(
        !is_significant ~ "Not Significant",
        local_i > 0 & .data[[variable]] > mean_val ~ "High-High",
        local_i > 0 & .data[[variable]] <= mean_val ~ "Low-Low",
        local_i < 0 & .data[[variable]] > mean_val ~ "High-Low",
        local_i < 0 & .data[[variable]] <= mean_val ~ "Low-High",
        TRUE ~ "Not Significant"
      )
    )
}

# Apply to alley light-outs
fishnet <- calculate_local_morans(fishnet, "countLights", k = 5)
```

```{r visualize-morans}
#| fig-width: 8
#| fig-height: 6

# Visualize hot spots
ggplot() +
  geom_sf(
    data = fishnet, 
    aes(fill = moran_class), 
    color = NA
  ) +
  scale_fill_manual(
    values = c(
      "High-High" = "#d7191c",
      "High-Low" = "#fdae61",
      "Low-High" = "#abd9e9",
      "Low-Low" = "#2c7bb6",
      "Not Significant" = "gray90"
    ),
    name = "Cluster Type"
  ) +
  labs(
    title = "Local Moran's I: Alley Light-Outs Clusters",
    subtitle = "High-High = Hot spots of disorder"
  ) +
  theme_crime()
```

Moran's I calculates hot spots and cold spots where there are a cluster of cells with high outage of alley lights and cluster of cells with low outage of alley lights. This allows us to more concretely determine significant hot spots, cold spots, and outliers.

This map shows clear hot spots scattered in an arc shape from north to south on the western side of Chicago. There are two primary cold spot, one large chunk in the southeast, and a long strip along the eastern side.

```{r distance-to-hotspots}
# Get centroids of "High-High" cells (hot spots)
hotspots <- fishnet %>%
  filter(moran_class == "High-High") %>%
  st_centroid()

# Calculate distance from each cell to nearest hot spot
if (nrow(hotspots) > 0) {
  fishnet <- fishnet %>%
    mutate(
      dist_to_hotspot = as.numeric(
        st_distance(st_centroid(fishnet), hotspots %>% st_union())
      )
    )
  
  cat("✓ Calculated distance to alley light-outs hot spots\n")
  cat("  - Number of hot spot cells:", nrow(hotspots), "\n")
} else {
  fishnet <- fishnet %>%
    mutate(dist_to_hotspot = 0)
  cat("⚠ No significant hot spots found\n")
}
```

Calculating each fishnet cell's center's distance to the closest hot spot.

```{r join-districts-&other-contexts}
# Join district information to fishnet
fishnet <- st_join(
  fishnet,
  policeDistricts,
  join = st_within,
  left = TRUE
) %>%
  filter(!is.na(District))  # Remove cells outside districts

cat("✓ Joined police districts\n")
cat("  - Districts:", length(unique(fishnet$District)), "\n")
cat("  - Cells:", nrow(fishnet), "\n")
```

#### Part 4: Count Regression Models

-   Fit Poisson regression
-   Fit Negative Binomial regression
-   Compare model fit (AIC)

```{r prepare-data}
# Create clean modeling dataset
fishnet_model <- fishnet %>%
  st_drop_geometry() %>%
  dplyr::select(
    uniqueID,
    District,
    countBurglaries,
    countLights,
    countVacant,
    lights_locations.nn,
    dist_to_hotspot
  ) %>%
  na.omit()  # Remove any remaining NAs

cat("✓ Prepared modeling data\n")
cat("  - Observations:", nrow(fishnet_model), "\n")
cat("  - Variables:", ncol(fishnet_model), "\n")
```

```{r fit-poisson}
# Fit Poisson regression
model_poisson <- glm(
  countBurglaries ~ countLights + lights_locations.nn + 
    dist_to_hotspot,
  data = fishnet_model,
  family = "poisson"
)

# Summary
summary(model_poisson)
```

Each additional broken light is associated with about a 0.15% increase in expected burglaries. A one-unit increase in nearest-neighbor broken light density is associated with about a 0.62% decrease in burglaries. Each additional meter of distance from a broken light hot spot reduces expected burglaries by about 0.006%. All effects are highly statistically significant with a p value of less than 0.001, indicating strong evidence that these relationships are not due to chance.

#### Part 5: Spatial Cross-Validation (2017)

-   Implement Leave-One-Group-Out cross-validation on 2017 data
-   Calculate and report error metrics (MAE, RMSE)

```{r spatial-cv}
# Get unique districts
districts <- unique(fishnet_model$District)
cv_results <- tibble()

cat("Running LOGO Cross-Validation...\n")

for (i in seq_along(districts)) {
  
  test_district <- districts[i]
  
  # Split data
  train_data <- fishnet_model %>% filter(District != test_district)
  test_data <- fishnet_model %>% filter(District == test_district)
  
  # Fit model on training data
  model_cv <- glm.nb(
    countBurglaries ~ countLights + lights_locations.nn + 
      dist_to_hotspot,
    data = train_data
  )
  
  # Predict on test data
  test_data <- test_data %>%
    mutate(
      prediction = predict(model_cv, test_data, type = "response")
    )
  
  # Calculate metrics
  mae <- mean(abs(test_data$countLights - test_data$prediction))
  rmse <- sqrt(mean((test_data$countLights - test_data$prediction)^2))
  
  # Store results
  cv_results <- bind_rows(
    cv_results,
    tibble(
      fold = i,
      test_district = test_district,
      n_test = nrow(test_data),
      mae = mae,
      rmse = rmse
    )
  )
  
  cat("  Fold", i, "/", length(districts), "- District", test_district, 
      "- MAE:", round(mae, 2), "\n")
}

# Overall results
cat("\n✓ Cross-Validation Complete\n")
cat("Mean MAE:", round(mean(cv_results$mae), 2), "\n")
cat("Mean RMSE:", round(mean(cv_results$rmse), 2), "\n")
```

```{r cv-results-table}
# Show results
cv_results %>%
  arrange(desc(mae)) %>%
  kable(
    digits = 2,
    caption = "LOGO CV Results by District"
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

Cross validations are performed to test the model and make sure that they model will hold under new and unforseen samples. After 22 folds, the results show that with a mean MAE of 97.57 and mean RMSE of 117.83, showing that my model's predictions are about 98 lights off the true prediction. The larger RMSE shows that there are some larger errors influencing the model.

#### Part 6: Model Evaluation

-   Compare to KDE baseline

```{r final-predictions}
# Fit final model on all data
final_model <- glm.nb(
  countBurglaries ~ countLights + lights_locations.nn + 
    dist_to_hotspot,
  data = fishnet_model
)

# Add predictions back to fishnet
fishnet <- fishnet %>%
  mutate(
    prediction_nb = predict(final_model, fishnet_model, type = "response")[match(uniqueID, fishnet_model$uniqueID)]
  )

# Also add KDE predictions (normalize to same scale as counts)
kde_sum <- sum(fishnet$kde_value, na.rm = TRUE)
count_sum <- sum(fishnet$countBurglaries, na.rm = TRUE)
fishnet <- fishnet %>%
  mutate(
    prediction_kde = (kde_value / kde_sum) * count_sum
  )
```

```{r compare-models}
#| fig-width: 12
#| fig-height: 4

# Create three maps
p1 <- ggplot() +
  geom_sf(data = fishnet, aes(fill = countBurglaries), color = NA) +
  scale_fill_viridis_c(name = "Count", option = "plasma", limits = c(0, 15)) +
  labs(title = "Actual Burglaries") +
  theme_crime()

p2 <- ggplot() +
  geom_sf(data = fishnet, aes(fill = prediction_nb), color = NA) +
  scale_fill_viridis_c(name = "Predicted", option = "plasma", limits = c(0, 15)) +
  labs(title = "Model Predictions (Neg. Binomial)") +
  theme_crime()

p3 <- ggplot() +
  geom_sf(data = fishnet, aes(fill = prediction_kde), color = NA) +
  scale_fill_viridis_c(name = "Predicted", option = "plasma", limits = c(0, 15)) +
  labs(title = "KDE Baseline Predictions") +
  theme_crime()

p1 + p2 + p3 +
  plot_annotation(
    title = "Actual vs. Predicted Burglaries",
    subtitle = "Does our complex model outperform simple KDE?"
  )
```

```{r model-comparison-metrics}
# Calculate performance metrics
comparison <- fishnet %>%
  st_drop_geometry() %>%
  filter(!is.na(prediction_nb), !is.na(prediction_kde)) %>%
  summarize(
    model_mae = mean(abs(countBurglaries - prediction_nb)),
    model_rmse = sqrt(mean((countBurglaries - prediction_nb)^2)),
    kde_mae = mean(abs(countBurglaries - prediction_kde)),
    kde_rmse = sqrt(mean((countBurglaries - prediction_kde)^2))
  )

comparison %>%
  pivot_longer(everything(), names_to = "metric", values_to = "value") %>%
  separate(metric, into = c("approach", "metric"), sep = "_") %>%
  pivot_wider(names_from = metric, values_from = value) %>%
  kable(
    digits = 2,
    caption = "Model Performance Comparison"
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

Overall, my model has a larger MAE and RMSE, but the difference is not too big, showing that my model is quite close to the baseline KDE prediction.
